Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-03-16 paper/rei.tex

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
\wh{Possibly add a few words on the PC version, list $\tau_a = \tau_a(z_{\rm min},z_{\rm max})$ in an equation or table, and refer to the
bottom panel of Fig.~\ref{fig:xe}}
The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.

@@

 \begin{figure}
          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau_prior_fac_0.8.pdf}
          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
           $\tau_{12}$ and \wh{range-bound flat - also if we change $\tau_{12}$ definitions change the numbers in the labels here} in $m_1-m_2$ (rotated rectangle, dashed lines) 
           %\ch{CH: [why is this flat in m1 and m2?]} 
           % flat in 2D or locally flat
           which are aligned with lines of constant $\tau_{12}$ (light gray lines). Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses). The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.} 
          \label{fig:prior_box}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{/paper/plots/plot_tau12_new_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{$\tau_{12,new}$ (no prior), with prior version is coming}
\label{fig:tau12}
\end{figure}

>>>>

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.   \wh{Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where
\begin{align}
    \tau_a(z,z_{\rm max}) &= 
    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{S_a(z) (1+z)^{2} }{H_{\rm fid}(z)}, \nonumber\\
    \tau_{\rm fid}(z,z_{\rm max}) &= 
    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e^{\rm fid}(z) (1+z)^{2} }{H_{\rm fid}(z)},
\end{equation}
 where ``fid" refers to evaluation in the fiducial cosmology.   When we omit the redshift arguments, we implicitly mean the total range, e.g. $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$. }

 \wh{
 In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
 Notice that positive values of the first component $m_1$ mainly provides the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.    Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
The underlying reason is that in $C_l^{EE}$ the higher the redshift, the
larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.  For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth.}

\wh{
In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a model like a near step function reionization and then extracts a single constraint on the total optical depth.}
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
\wh{Here we take $x_e^{\rm max} = 1+ n_{\rm He}/n_{\rm H}$ to account for the contribution of singly ionized helium.}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.

@@

 \begin{figure}
          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau12.pdf}
          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
           $\tau_{12}$ and \wh{range-bound flat - also if we change $\tau_{12}$ definitions change the numbers in the labels here} in $m_1-m_2$ (rotated rectangle, dashed lines) 
           %\ch{CH: [why is this flat in m1 and m2?]} 
           % flat in 2D or locally flat
           which are aligned with lines of constant $\tau_{12}$ (light gray lines). Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses). The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.} 
          \label{fig:prior_box}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf
}
\caption{Internal (with prior version)}
\label{fig:tau12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
\caption{Internal (with prior version)}
\label{fig:tau12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{/project2/whu5/chenhe/git/ReiPl18-Results/paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
\caption{$\tau_{12,new}$ \ch{(took out the cosmology part, only shrunk all posteriors very slightly and shifted them a bit to the left ); with prior version now added, it uses the mean from m3, m4, m5 that are computed from the chains with prior as well.}}
\label{fig:tau12}
\end{figure}
